---
title: "Geospatial Workshop PPSP 2024"
author: "Afiq Amsyar"
format:
  html:
    toc: true
    toc-location: left
    theme: 
      light: cosmo
      dark: 
        - cosmo
        - theme-dark.scss
    grid:
      sidebar-width: 300px
      margin-width: 300px
      body-width: 900px
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
  docx: 
    toc: true
    toc-depth: 2
self-contained: true
---

## Spatial Packages in R

Ensure the following R packages are installed beforehand.

```{r, eval = FALSE}
install.packages(c("sf", "terra", "geodata", "rnaturalearth", "spdep",

                   "dplyr", "SpatialEpi", "wbstats", "flexdashboard",

                   "ggplot2", "viridis", "RColorBrewer", "patchwork", "DT",

                   "leaflet", "mapview", "leafpop", "leafsync", "rasterVis"))
```

```{r, eval = FALSE}
install.packages("INLA",
repos = "https://inla.r-inla-download.org/R/stable", dep = TRUE)
```

## Load Packages

For this session, we will load some of the packages needed for plotting basic spatial data

```{r}
library(sf)
library(mapview)
library(janitor)
library(readxl)
library(dplyr)
library(ggplot2)
library(DT)
```

## Making Maps with R

```{r}
kelantan <- st_read("kelantan.shp")
```

### Plot 1

```{r}
mapview(kelantan[,3])
```

### Plot 2

```{r}
# Create a custom popup with multiple fields
popup_info <- paste0(
  "<strong>MUKIM: </strong>", kelantan$MUKIM, "<br>",
  "<strong>DAERAH: </strong>", kelantan$DAERAH, "<br>",
  "<strong>Population: </strong>", kelantan$JUM_JANTIN
)
mapView(kelantan, zcol = "JUM_JANTIN", layer.name = "Population", popup = popup_info)
```

## Adding Spatial Attributes

### Leptospirosis Data

```{r}
lepto <- read_xlsx("leptospirosis.xlsx") %>% clean_names()
glimpse(lepto)
```

### **Convert Leptospirosis data to spatial data**

Use st_as_sf ( ) function to convert line listing data that contained Lat/Long attributes to spatial object

```{r}
lepto <- st_as_sf(lepto, 
                    coords = c("longitude_wgs", "latitude_wgs"), 
                    crs = 4326)
```

Confirm CRS is wgs84

```{r}
st_crs(lepto)
```

### **Convert shapefile to RSO to match with Kelantan Map**

```{r}
lepto_2 <- st_transform(lepto, 3168)
```

### **Plot map to see outlier**

```{r}
ggplot() +
  geom_sf(data = lepto_2) +
  ggtitle("Map of Leptospirosis") +
  theme_bw()
```

### Plot the cases over the Kelantan map

```{r}
overall_plot <- ggplot() +
  geom_sf(data = kelantan) +   #base map
  geom_sf(data = lepto_2, color = "red", size = 0.5) +  #cases in spatial data
  ggtitle("Map of Leptospirosis Cases in Kelantan for 2016-2022") +
  theme_bw() +  
  theme(plot.title = element_text(size = 12),  strip.text = element_text(size = 12)) # cosmetic

overall_plot 
```

```{r}
overall_plot + 
  facet_wrap(~tahun_daftar) + #to plot according to year
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5)) +
  theme(plot.title = element_text(size = 12),  strip.text = element_text(size = 12)) +
  ggtitle("Map of Leptospirosis Cases in Kelantan for 2016-2022")
```

### **Using OSM as basemap**

```{r}
library(leaflet)

leaflet(lepto) |>
  addTiles() |>  # Add default OpenStreetMap tiles
  addCircleMarkers(
    radius = 2,   # Adjust marker size
    color = "red",  # Marker color
    stroke = FALSE, fillOpacity = 0.8)
```

## **Aggregated Data**

For this exercise, we focus our analysis on Leptospirosis cases reported in 2016 only.

```{r}
lepto_16 <- lepto_2 %>% filter(tahun_daftar == "2016")
```

### **Joint point to polygon**

```{r}
st_crs(lepto_16)
```

### **Count all leptospirosis cases for each mukim in 2016**

```{r}
count_lepto_sub_yr <- lepto_16 %>% 
  count(daerah, mukim) %>% 
  ungroup()
```

### **Joining the count data to base-map of Kelantan**

```{r}
kelantan_count_lepto <- st_join(kelantan, count_lepto_sub_yr)
```

### Map the count of cases according to subdistrict in Kelantan

```{r}
# Create a custom popup with multiple fields
popup_info <- paste0(
  "<strong>MUKIM: </strong>", kelantan_count_lepto$MUKIM, "<br>",
  "<strong>DAERAH: </strong>", kelantan_count_lepto$DAERAH, "<br>",
  "<strong>Leptospirosis Cases: </strong>", kelantan_count_lepto$n
)

# Display the map with the custom popup
mapView(kelantan_count_lepto, zcol = "n", layer.name = "Count of Leptospirosis cases 2016", popup = popup_info)
```
